---
title: "Frequencies and Histograms"
author: "Gaston Sanchez"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NBA Data Set

```{r eval = FALSE}
# download data set into your working directory
url <- "https://raw.githubusercontent.com/ucb-introstat/introstat-spring-2019/master/data/nba_players.csv"
dat <- read.csv(url)
```

```{r echo = FALSE}
dat <- read.csv('../data/nba_players.csv')
```

The dataset `dat` is a `data.frame` (a tabular object) with dimensions (i.e. number of rows and columns):

```{r}
dim(dat)
```

In turn, the available variables (in the columns) are:

```{r}
colnames(dat)
```

To take a look at a particular variable, you can type the name of the table, followed by the `$` sign, followed by the name of the variable:

```{r eval = FALSE}
dat$player
dat$position
```


## Frequency Tables

A common first step in examining the values of a variable is to count the frequencies of such values. This can be easily done in R with the function `table()`. For instance, to get the counts of the different types of positions, run the following command:

```{r}
# position frequencies
table(dat$position)
```


```{r}
# age frequencies
table(dat$age)
```


### Relative Frequencies

In order to get __relative frequencies__ (i.e. proportions or fractions) you need to divide the frequencies by the total number of observations. You can do this in R in several ways:

```{r}
# position relative frequencies (option 1)
table(dat$position) / nrow(dat)

# position relative frequencies (option 2)
table(dat$position) / length(dat$position)

# position relative frequencies (option 3)
prop.table(table(dat$position))
```


## Bar-Charts and Pie-Charts

To visualize frequencies (or relative frequencies) you can create a bar-chart with the function `barplot()`. All you need is to take a `table()` and pass it inside `barplot()`:

```{r barplot1}
barplot(table(dat$position))
```

Another common graph to visualize frequencies is a pie-chart, which can be produced in R with the function `pie()`. Like in a bar-chart, you can pass a `table()` to `pie()`:

```{r piechart1}
pie(table(dat$position))
```


## Categorizing Quantitative Variables

Let's consider the quantitative variable `height`, obtaining frequencies and a bar-chart. 

```{r}
table(dat$height)
```

```{r barplot2}
barplot(table(dat$height), las = 1, border = NA)
```

The bar-chart reveals some information but there are some issues. Can you guess what are those issues?

Or consider `weight`. Things can get a bit overwhelming:

```{r}
table(dat$weight)
```

```{r barplot3}
barplot(table(dat$weight))
```

In these cases, it's better to reorganize the values into groups or categories. For example, we could form groups 160-180 pounds, 180-200, 200-220, etc:

```{r}
weight2 <- cut(
  dat$weight, 
  breaks = seq(from = 160, to = 300, by = 20))

table(weight2)
```

and then make a bar-chart of the categorized variable `weight2`

```{r barplot4}
barplot(table(weight2), las = 1, border = NA)
```


## Histograms

In addition to using bar-charts to visualize the frequencies of a quantitative variable, we can also use a __histogram__.

R has the `hist()` function that allows you to create a basic histogram:

```{r hist1}
# default histogram
hist(dat$weight)
```

We can specify some graphical parameters to customize the appearence of the histogram:

```{r hist2}
hist(dat$weight, las = 1, col = 'gray80')
```

More interestingly, we can also play with the argument `breaks` to change the number of bars:

```{r hist3}
hist(dat$weight, las = 1, col = 'gray80', breaks = 8)
```

Here's another option:

```{r hist4}
hist(dat$weight, las = 1, col = 'gray80', 
     breaks = seq(from = 160, to = 310, by = 10))
```

And one more histogram with more "resolution":

```{r hist5}
hist(dat$weight, las = 1, col = 'gray80', 
     breaks = seq(from = 160, to = 310, by = 5))
```

